**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
draft of all SME revisions completed.

In this study, we focus on the state of Rhode Island for several reasons.  Rhode Island has the second highest population density in the country, yet still has a mix of land use/land cover with developed areas making up ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Development"]`% of the state, forested areas ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Forest"]`%, wetlands ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Wetlands"]`%, and agriculture ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Planted/Cultivated"]`%.  Given these facts, Rhode Island lakes occur across a gradient of land use/land cover and face a suite of human pressures that most lakes in the nation are also facing with nutrients and pathogens of particular concern in the state [@ri_2016_waterquality].  Lastly, Rhode Island provides an ideal study area as significant, long-term datasets are available for lakes making it possible to examine water quality trends.

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R packages and operating system used for this analysis are included in a file, `sessioninfo.txt` at <>[].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].

@@

1. For each site, calculate the annual medians, producing a single median value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site medians across years are assumed to be independent.
2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
3. Calculate the anomaly for each annual median at each site by subtracting the sites reference long-term medians from the sites yearly median.
4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

After filtering and summarizing the data some years may not have sufficient number of sites to be included.  We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data (`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).

@@

For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.  A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero.

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

Median annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with some, albeit weak, statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.  Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Twenty year trend for median TN anomaly.  Panel A. URI Watershed Watch yearly TN anomalies.  Panel B. LAGOSNE yearly TN anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A.). Also, 1998 only had three sites with available total nitrogen data and may not be representative.  This year also recorded the lowest median value.  Thus, 1998 may be skewing these results.  When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.  Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus shows essentially no trend over time in the yearly anomalies with little statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Twenty year trend for median TP anomaly.  Panel A. URI Watershed Watch yearly TP anomalies.  Panel B. LAGOSNE yearly TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.  The TN:TP ratio has a small slope, very little statistical support, no pattern in the above and below years, and a slope magnitude less than 0.26 (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Twenty year trend for median TN:TP anomaly.  Panel A. URI Watershed Watch yearly TN:TP anomalies.  Panel B. LAGOSNE yearly TN:TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In addition to the sub-regional annualized trends of the five variables we address with this study, other trends may also be of interest.  For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms.  Many national and regional studies have documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].  Evidence in Rhode Island also points to possible lengthening of the growing season as both May and October, the beginning and end of the sampling dates in our dataset, show increasing temperature trends over time [@noaatrends].  Furthermore, preliminary analysis of the URIWW data supports the idea that growing seasons may be getting longer in Rhode Island.  Cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].  Early in our study period (1993-1995), the average first day of lake temperatures exceeding 25°C was June 11th, at the end of our study period (2014-2016) these warmer temperature were seen, on average, on June 1st.  A site-specific anomaly analysis of growing season length could shed further light on potential changes to growing seasons in Rhode Island lakes.

Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.  We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.  This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.  Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only a single rate for all lakes.  Furthermore, Stoddard et al. [-@stoddard2016continental] see an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.  A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al. [-@stoddard2016continental], that might be occuring only on oligotrophic lakes.   A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.

@@

Our analysis has shown warming lakes in Rhode Island, but broad scale warming of lakes has also been documented and shown to be a result of both climatic and local drivers and can vary greatly within regions [e.g., @oreilly2015warming].  Given that the drivers of warming are at both broad and local scales, suggests that managing warming lake temperatures will be a difficult task.  To counteract the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions. Reducing nutrient loads below current levels is one such intervention and could be acheieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  In any event, controlling future eutrophication and protecting against harmful aglal blooms will continue to be a challenge against the backdrop of warming lakes.

@@

```{r year_num_table_ww, echo=FALSE, results="asis"}
#pander::pander(year_num_ww, format = "pipe")
knitr::kable(year_num_ww, "pipe")
```
`r tab_num("year_n_ww")`

```{r year_num_table_lagos, echo=FALSE, results="asis"}
knitr::kable(year_num_lagosne, format = "pipe")
```
`r tab_num("year_n_lagosne")`

@@

@@

![](../../figures/simulated_data.jpg)
`r fig_num("simulated_data")`

![](../../figures/simulated_analysis.jpg)
`r fig_num("simulated_analysis")`

>>>>

In this study, we focus on the state of Rhode Island for several reasons.  Rhode Island has the second highest population density in the country, yet still has a mix of land use/land cover with developed areas making up ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Development"]`% of the state, forested areas ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Forest"]`%, wetlands ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Wetlands"]`%, and agriculture ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Planted/Cultivated"]`%. The remainder of the land area is made up other land use/land cover classes.  Given these facts, Rhode Island lakes occur across a gradient of land use/land cover and face a suite of human pressures that most lakes in the nation are also facing with nutrients and pathogens being of particular concern in the state [@ri_2016_waterquality].  Lastly, Rhode Island provides an ideal study area as significant, long-term datasets are available for lakes making it possible to examine water quality trends.

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.  For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].

@@

1. For each site, calculate the annual medians, producing a single median value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site medians across years are assumed to be independent.
2. Calculate the long-term reference median for each site. This results in a single site-specific long-term median.
3. Calculate the anomaly for each annual median at each site by subtracting the sites reference long-term medians from the sites yearly median.
4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

After filtering and summarizing the data some years may not have sufficient number of sites to be included.  We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data and late in the time frame for the LAGOSNE data (`r fig_num("year_n", display = "cite", caption = "Number of sites available for trend analysis by parameter and year for both URIWW and LAGOSNE.")`).

@@

For changes in chlorophyll, total phosphorus, and total nitrogen, we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.  A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero.

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorophyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

Median annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with some, albeit weak, statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.  Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Twenty year trend for median TN anomaly.  Panel A. URI Watershed Watch yearly TN anomalies.  Panel B. LAGOSNE yearly TN anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.  Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A.). Also, 1998 only had three sites with available total nitrogen data and may not be representative.  This year also recorded the lowest median value.  Thus, 1998 may be skewing these results.  When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.  Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus shows essentially no trend over time in the yearly anomalies with little statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Twenty year trend for median TP anomaly.  Panel A. URI Watershed Watch yearly TP anomalies.  Panel B. LAGOSNE yearly TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.  The TN:TP ratio has a small slope, very little statistical support, no pattern in the above and below years, and a slope magnitude less than 0.26 (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Twenty year trend for median TN:TP anomaly.  Panel A. URI Watershed Watch yearly TN:TP anomalies.  Panel B. LAGOSNE yearly TN:TP anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.  Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.  Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In addition to the sub-regional annualized trends of the five variables we address with this study, other trends may also be of interest.  For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms as cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].  Many national and regional studies have also documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].  Evidence in Rhode Island also points to possible lengthening of the growing season as both May and October, the beginning and end of the sampling dates in our dataset, show increasing temperature trends over time [@noaatrends].  Furthermore, preliminary analysis of the URIWW data supports the idea that growing seasons may be getting longer in Rhode Island.    Early in our study period (1993-1995), the average first day of lake temperatures exceeding 25°C was June 11th, at the end of our study period (2014-2016) these warmer temperature were seen, on average, on June 1st.  A site-specific anomaly analysis of growing season length could shed further light on potential changes to growing seasons in Rhode Island lakes.

Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.  We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.  This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.  Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only on a single rate for all lakes.  Furthermore, Stoddard et al. [-@stoddard2016continental] report an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.  A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al. [-@stoddard2016continental], that might be occuring only on oligotrophic lakes.   A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.

@@

Our analysis has shown warming lakes in Rhode Island, but broad scale warming of lakes has also been documented and shown to be a result of both climatic and local drivers and can vary greatly within regions [e.g., @oreilly2015warming].  Given that the drivers of warming are at both broad and local scales, managing warming lake temperatures will be a difficult task.  Counteracting the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions. Reducing nutrient loads below current levels is one such intervention and could be achieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  In any event, controlling future eutrophication and protecting against harmful aglal blooms will continue to be a challenge against the backdrop of warming lakes.

@@

@@

![](../../figures/sample_sizes.jpg)

`r fig_num("year_n")`

@@

![](../../figures/simulated_data_analysis.jpg)
`r fig_num("simulated")`
